---
title: "Evaluating Physicochemical Drivers of River Flow Using Weighted Regression"
author: "Kenshi King"
date: "2025-11-28"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Data Set Credits

Author: Daniel Fedorov

Source: Kaggle

```{r}
#imports 
library(tidyverse) #data wrangling
library(ggplot2) #visualizations
library(ggcorrplot) #correlation plots
library(lmtest) #Breusch-Pagan test
library(car) #vif, Box-Cox test, components + residuals plot
library(caTools) #train/test split
library(broom) #coefficient plot
```

```{r}
sessionInfo()
```

```{r}
list.files("figures")
dir.create("figures", showWarnings = FALSE)
```

## Key

This key contains relevant variables to our analysis and their units of
measurement:

1\. Average Water Speed (cm/s)

2\. Average Water Direction (degrees)

3\. Chlorophyll (ug/L)

4\. Temperature (°C)

5\. Dissolved Oxygen (mg/L)

6\. Dissolved Oxygen (% Saturation)

7\. pH

8\. Salinity (ppm)

9\. Specific Conductance (mS/cm)

10\. Turbidity (Formazin Nephelometric Units (FNU))

## Introduction

This project examines a dataset provided by the Queensland Government
Open Data Portal containing measurements of water physicochemcical
parameters of the Brisbane River in Australia (Fedorov). The dataset
includes timestamps, record numbers, and various characteristics of the
river during a given instance. Using this data, we aim to test whether
or not the average water speed is influenced by these river
characteristics. Understanding water speed is crucial because it affects
numerous things: river habitats and aquatic biodiversity, agriculture,
hydroelectric power generation, erosion, flood risk management, and many
others. Monitoring these speeds is tantamount to ensuring risk
management and environmental protection (Fain).

```{r}
#Loading in our dataset
df <- read_csv('brisbane_water_quality.csv')
head(df)
```

## Section 1: Descriptive Statistics and Checking Assumptions 

We begin our analysis by first examining some characteristics of the
dataset. Descriptive statistics are helpful in arming us with useful
data points as well as allowing us to detect potential outliers whether
that be highly correlated variables or null values (NA). We will ignore
columns Timestamp column as it may over complicate our analysis and the
various [quality] columns due to their long stretch of null values. 

```{r}
water <- df %>% 
  select(`Average Water Speed`, `Average Water Direction`, Chlorophyll, Temperature, `Dissolved Oxygen`, pH, Salinity, `Specific Conductance`, Turbidity)
water
```

This dataset contains over 30,000 observations of the Brisbane River,
capturing variation in the water's physicochemical parameters across
30-minute intervals. Before cleaning, many variables contained a
substantial amount of missing values, most notably temperature (5164
missing), dissolved oxygen (4300 missing), salinity (3958 missing), and
turbidity (2000 missing). After removing these rows to ensure we only
have complete observations, the dataset retained the majority of its
observations while dropping potential inconsistencies. (Potentially
revise to fill the missing with avg or median values, depends on the gap
lengths\*)

```{r}
clean_water <- na.omit(water)
```

```{r}
#Uncleaned and cleaned placeholders
water_plot <- water %>% 
  select(`Average Water Speed`) %>% 
           mutate(stage = "Before Cleaning")
clean_water_plot <- clean_water %>% 
  select(`Average Water Speed`) %>% 
  mutate(stage = "After Cleaning")

combined <- bind_rows(water_plot, clean_water_plot)

combined$stage <- factor(combined$stage, levels = c("Before Cleaning", "After Cleaning"))

ggplot(combined, aes(x = `Average Water Speed`)) + 
  geom_histogram(binwidth=2, fill="blue", color="white") + 
  facet_wrap(~stage) +
  labs(title = "Distribtuon of Average",
       x = NULL,
       y = "Average Water Speed (cm/s)") +
  theme_minimal() +
  theme(legend.position = "none")
ggsave("figures/water_speed_distribution.png", 
       plot = last_plot(),  # or replace with your saved plot object, e.g. speed_plot
       width = 8, height = 5, dpi = 300)
```

Comparing the summary statistics before and after cleaning the data,
some key trends slightly shifted, reflecting the removal of those
null/extreme values. For instance, the mean average water speed
decreased from \~16.80 to \~16.01, and the median from \~13.39 to
\~12.33; the range remained wide, indicating some persistent extreme
values. With regards to the other average water direction and the
physicochemical parameters, they also exhibited minor changes. (Weighted
regression will allow us to skip z-score scaling)

Overall, the removal of these missing values slightly adjusted the
descriptive statistics, retaining variability and the general patterns
of the dataset. This version provides us with a more reliable starting
point for further analysis.

```{r}
summary(water)
```

```{r}
summary(clean_water)
```

## Section 2: Model Preparation and Fitting

Here, we are interested in seeing if there are any variables with
relationships that are a bit too strong. Highly correlated pairs can
lead to multicollinearity, which can inflate standard errors and lower
our model's estimates and predicting power. On a 0 to 1 scale, a common
"rule-of-thumb" threshold is usually \<0.7, with any pairs going past
this potentially indicating multicollinearity. As shown below, Salinity
and Specific Conductance are extremely correlated (\~0.999), which makes
sense logically as conductance of water is highly dependent on its salt
contents so if a river is measured to have a high amount of salinity,
its specific conductance will also be high. To address this, we will
remove Specific Conductance from our model, retaining Salinity as its
presenter. 

```{r}
cor(clean_water[,-c(1,9)])
```

```{r}
predictors <- clean_water %>% 
  select(`Average Water Direction`, Chlorophyll, Temperature, `Dissolved Oxygen`, pH, Salinity, `Specific Conductance`, Turbidity)
corr_matrix <- cor(predictors)

corr_plot <- ggcorrplot(corr_matrix, method = "square",
           type = "lower",
           lab = TRUE,
           lab_size = 3,
           colors = c("red", "white", "steelblue"),
           title = "Correlation Matrix of Physicochemical Predictors",
           ggtheme = theme_minimal()) + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  theme(plot.title = element_text(hjust = 0.65))
corr_plot
ggsave("figures/correlation_heatmap_before.png", plot = corr_plot, width = 7, height = 6, dpi = 300) 
```

After we removed Specific Conductance, no pair of variables exceeded the
0.7 threshold anymore. 

```{r}
cor(clean_water[,c(1,8)])
```

```{r}
predictors_cleaned <- clean_water %>% 
  select(`Average Water Direction`, Chlorophyll, Temperature, `Dissolved Oxygen`, pH, Salinity, Turbidity)
corr_matrix <- cor(predictors_cleaned)

corr_plot <- ggcorrplot(corr_matrix, method = "square",
           type = "lower",
           lab = TRUE,
           lab_size = 3,
           colors = c("red", "white", "steelblue"),
           title = "Correlation Matrix of Physicochemical Predictors (Omitted Conductance)",
           ggtheme = theme_minimal()) +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.65))
corr_plot
ggsave("figures/correlation_heatmap_after.png", 
       plot = corr_plot, 
       width = 7, height = 6, dpi = 300)
```

We create a multiple linear regression model with Average Water Speed
being our response variable and the remaining variables (Average Water
Direction, Chlorophyll, Temperature, Dissolved Oxygen, pH, Salinity,
Turbidity) as our predictors. We will approach this using manual
selection, later using step-wise selection to assess if they align. 

Looking at the output, we notice that the p-values for Average Water
Direction (0.0971) and Temperature (p = 0.5361) exceeds the 0.05
significance threshold, indicating that they may not significantly
affect/predict average water direction; we will consider removing
temperature first and refitting.

```{r}
model <- lm(`Average Water Speed` ~ `Average Water Direction` + Chlorophyll+Temperature+`Dissolved Oxygen`+pH+Salinity+Turbidity, data=clean_water) 
summary(model) 
```

Upon refitting, Average Water Direction's p-value (0.0893) still exceeds
the 0.05 threshold so we will remove it and refit once more.

```{r}
model2 <- lm(`Average Water Speed` ~ `Average Water Direction` + Chlorophyll+`Dissolved Oxygen`+pH+Salinity+Turbidity, data=clean_water) 
summary(model2)
```

The tentative final model includes Chlorophyll, Dissolved Oxygen, pH,
Salinity, and Turbidity. Temperature and Average Water Direction were
excluded due to lack of significance. When using stepwise selection
using Bayesian Information Criterion (BIC) penalty (since it scales with
sample size), it resulted in the same model, indicating that this model
provides a balance between goodness of fit and model complexity. 

The adjusted R\^2 of 0.137 indicates that our model explains a moderate
portion of variation in average water speed. Much of the remaining
variability is likely driven by environmental factors not captured in
this dataset. Regardless, our predictors' statistical significance do
prove that they truly have an effect on determining average water speed,
even if the model doesn't explain most of the variation. Rivers are
inherently dynamic, as their flow can be strongly affected by a variety
of external forces and conditions (Trent and Brown).

```{r}
full_model <- lm(`Average Water Speed`~Chlorophyll+`Dissolved Oxygen`+pH+Salinity+Turbidity, data=clean_water)
summary(full_model)
```

```{r}
step(model, direction="both", k=log(20028))
```

Furthermore, the Variance Inflation Factor (VIF) values for all
predictors are well below the common threshold of 5, indicating that
multicollinearity is not a concern of this model.

```{r}
vif(full_model)
```

## Section 3: Diagnostics

In order to ensure the assumptions of our linear regression model are
met, several diagnostic checks were performed. We are looking to confirm
if the assumptions of linearity, constant variance, and normality of
residuals are satisfied. The fitted values versus residuals was plotted
to visually show if any patterns are recognized and we see that there is
this downward, fanning of the point suggesting possible
heteroscedasticity. The Breusch-Pagan Test (BP) yielded an extremely
small p- value (\<2.2e-16), undeniably rejecting the null hypothesis of
constant variance, confirming our suspicion of heteroscedasticity. 

```{r}
resid_df <- data.frame(fitted = full_model$fitted.values,
                       residuals = full_model$residuals)

resid_plot <- ggplot(resid_df, aes(x=fitted, y=residuals)) +
  geom_point(alpha=0.4, color="steelblue") +
  geom_hline(yintercept=0, color="firebrick", linetype="dashed") +
  labs(title="Residuals vs. Fitted Values",
       x="Fitted Values",
       y="Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
full_model$residuals
resid_plot

ggsave("figures/residuals_vs_fitted.png", 
       plot = resid_plot, 
       width = 7, height = 6, dpi = 300)
```

Our Q-Q plot shows a slightly exponential trend beginning to form about
x-axis = 1, causing some deviation in our linearity assumption. Our
Component + Residuals (crPlots) also agree with this deviation of
linearity, indicating that some transformation may be necessary. 

```{r}
qq_plot <- ggplot(resid_df, aes(sample=residuals)) +
    stat_qq(color="steelblue", alpha=0.5) +
    stat_qq_line(color="firebrick", linetype="dashed") +
    labs(title="Normal Q-Q Plot",
         x="Theoretical Quantiles",
         y="Sample Quantiles") +
    theme_minimal() +
    theme(plot.title = element_text(hjust=0.5))
qq_plot
ggsave("figures/qq_plot.png", 
       plot = qq_plot, 
       width = 7, height = 6, dpi = 300)
```

```{r}
crPlots(full_model)
cook <- cooks.distance(full_model)
max(cook)
```

A Box-Cox transformation was performed to determine whether a power
transformation to the response would improve model assumptions of
linearity. The analysis suggested that the optimal lambda value is 1,
indicating that no transformation is necessary. Therefore, the response
variable is retained on its original scale.

```{r}
box_cox <- powerTransform(`Average Water Speed` ~Chlorophyll+`Dissolved Oxygen` +pH+Salinity+Turbidity, data=clean_water) 
summary(box_cox) 
```

To address the issue of heteroscedasticity, we apply the Weighted Least
Squares (WLS) to give less influence to observations with higher
variance in residuals, resulting in more reliable coefficient estimates
and improving the homoscedasticity assumption. 

```{r}
ols <- lm(`Average Water Speed` ~Chlorophyll+`Dissolved Oxygen` +pH+Salinity+Turbidity, data=clean_water)
residual_sq <- resid(ols)^2
var_model <- lm(residual_sq~fitted(ols))
pred_var <- predict(var_model)
weight <- (1/pred_var)
wls <- lm(`Average Water Speed` ~Chlorophyll+`Dissolved Oxygen` +pH+Salinity+Turbidity, data=clean_water, weights=weight)
summary(wls)
```

Since applying the WLS approach, the BP test yielded a higher p-value
(0.000204), indicating that while some heteroscedasticity remains, the
variance of residuals has improved compared to the original model. 

```{r}
bptest(wls)
```

Using the outlier test, we identified 10 observations with unusually
high average water speeds that are extreme compared to the rest of the
dataset. These values have the potential to unduly influence the
regression results so we will remove them in an effort to improve model
reliability. 

```{r}
outlierTest(wls)
```

```{r}
outlier <- c(1,2,3340,3805,4104,3385,3524,3746,3938,5051,11844)
clean_water[outlier,] 
```

```{r}
final <- clean_water[!(seq_len(nrow(clean_water)) %in% outlier), ]
final
```

```{r}
final_ols <- lm(`Average Water Speed` ~Chlorophyll+`Dissolved Oxygen` +pH+Salinity+Turbidity, data=final)
residual_sq <- resid(final_ols)^2
var_model<-lm(residual_sq~fitted(final_ols))
pred_var<- predict(var_model)
final_weight <- (1/pred_var)
```

After refitting the final model, including the weight, the model still
exhibits heteroscedasticity as it still fails to reject the BP test's
null hypothesis. However, the weighted residuals indicate that
heteroscedasticity has been reduced once more, reflected by the higher
p-value (p \~ 0.00011) compared to the original (\<2.2e-16) and prior
(0. 

```{r}
final_wls <- lm(`Average Water Speed` ~Chlorophyll+`Dissolved Oxygen` +pH+Salinity+Turbidity, data=final, weights=final_weight)
bptest(final_wls)
```

## Section 4: Predictive Validation

In order to assess whether or not the model generalizes beyond the data
it was fit on, we implement a train/test split approach into a 80% and
20% subset, respectively. The weighted least squares model was refit
using only the training data, with weights recalculated with training
residuals to avoid overlap with the test set. The fitted model was then
used to predict Average Water Speed.

```{r}
#randomizing seed for reproducibility + 7 is lucky
set.seed(777)

#Split
split <- sample.split(final$`Average Water Speed`, SplitRatio = 0.8)

#Train/Test sets
training_set <- subset(final, split == TRUE)
test_set <- subset(final, split == FALSE)

#Refit WLS on training data only
train_ols <- lm(`Average Water Speed`~Chlorophyll+`Dissolved Oxygen`+pH+Salinity+Turbidity, data=training_set)
train_resid_sq <- resid(train_ols)^2
train_var_model <- lm(train_resid_sq~fitted(train_ols))
train_pred_var <- predict(train_var_model)
train_weight <- 1/train_pred_var
train_wls <- lm(`Average Water Speed`~Chlorophyll+`Dissolved Oxygen`+pH+Salinity+Turbidity, data=training_set, weights=train_weight)
summary(train_wls)

#Test 
test_pred <- predict(train_wls, newdata=test_set)

#Results
test_actual <- test_set$`Average Water Speed`
rmse <- sqrt(mean((test_pred-test_actual)^2))
mae <- mean(abs(test_pred-test_actual))

rmse
mae
```

The resulting test set RMSE was 12.02 cm/s and MAE was 8.98 cm/s. The
RMSE aligns closely with the residual standard error of the original
unweighted model (\~12.08 cm/s), suggesting that the model's predictive
performance on unseen data is consistent with its in sample fit rather
than reflecting overfitting.

```{r}
pred_df <- data.frame(actual = test_set$`Average Water Speed`,
                      predicted = test_pred)

pred_plot <- ggplot(pred_df, aes(x=actual, y=predicted)) +
  geom_point(alpha=0.4, color="steelblue") +
  geom_abline(slope=1, intercept=0, color="firebrick", linetype="dashed") +
  labs(title="Predicted vs. Actual Average Water Speed (Test Set)", 
       x="Actual Average Water Speed (cm/s)",
       y="Predicted Average Water Speed (cm/s)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
pred_plot
ggsave("figures/predicted_vs_actual.png", 
       plot = pred_plot, 
       width = 7, height = 6, dpi = 300)
```

## Section 5: Analysis

The final weighted least squares model depicts how Chlorophyll,
Dissolved Oxygen, pH, Salinity, and Turbidity influence the average
water speed of the Brisbane River. Within the parameter estimates, all
coefficients are statistically significant at the 0.01 level, indicating
well-supported points. Notably, dissolved oxygen has a strong negative
effect (Estimate \~ -4.91), suggesting that higher oxygen levels are
correlated with lower water speeds. pH displays a very strong, positive
effect (Estimate 37.73), implying higher pH associates with faster water
speeds. Salinity has a moderately negative effect (Estimate \~ -0.97),
while Turbidity (Estimate\~ 0.25) and Chlorophyll (Estimate\~ -0.15)
produced smaller but still significant effects. The model explains
roughly 17.4% of variability in water speed, suggesting that other
environmental factors not included within the dataset likely contribute
to variation. 

```{r}
summary(final_wls)
```

```{r}
coef_df <- tidy(final_wls, conf.int = TRUE) %>%
  filter(term !="(Intercept)")

coef_plot <- ggplot(coef_df, aes(x = estimate, y = reorder(term, estimate))) +
  geom_point(color = "steelblue", size = 3) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), 
                width = 0.2, color = "steelblue", orientation = "y") +
  geom_vline(xintercept = 0, color = "firebrick", linetype = "dashed") +
  labs(title = "Coefficient Estimates with 95% Confidence Intervals",
       x = "Estimate",
       y = NULL) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
coef_plot
ggsave("figures/coefficient_plot.png", 
       plot = coef_plot, 
       width = 7, height = 5, dpi = 300)
```

```{r}
plot(final_wls$fitted.values, final_wls$residuals)
abline(h=0, col="red")
```

## Section 6: Limitations and Future Implementations

While the final model provides statistically significant and
interpretable relationships between physiochemical parameters and
average water speed, several limitations should be acknowledged. First,
the model explains only a modest portion of the variability in water
speed (Adjusted R2 ≈ 0.174), suggesting that important drivers of river
flow are not captured in this dataset. According to the research (Less
et al. 1181), hydrological factors such as rainfall, tidal cycles, and
upstream discharge volume are known to strongly influence river
velocity. However, these were unavailable in the source data, and their
absence likely accounts for much of the unexplained variance.

Second, the time stamp variable was excluded from modeling to simplify
the analysis, meaning temporal structure such as diurnal tidal patterns
or seasonal variation, was not directly incorporated. Given that river
flow is inherent interdependent, future work could explore this
structure directly, either through time-series methods that account for
autocorrelation between sequential 30-minute readings, or through
mixed-effects models that treat time-of-day or season as structured
random effects.

Finally, the train/test validation performed did give us confirmation
that the data generalizes well with unseen data, but this could be
further extended using k-fold cross-validation to give us a greater
range of estimates across many partitions to better defend our findings.

## Conclusion

This analysis provides a comprehensive assessment of how key
physicochemical parameters influence the average water speed of the
Brisbane River. Using multiple linear regression and rigidly considering
model assumptions, we identified Chlorophyll, Dissolved Oxygen, pH,
Salinity, and Turbidity as significant predictors of water speed. The
Weighted Least Squares approach improved the reliability of coefficient
estimates by reducing the influence of heteroscedasticity, and removal
of extreme outliers further strengthened model stability. While the
final model explains approximately 17.4% of the variability in water
speed, the statistical significance of all predictors demonstrates that
these environmental factors meaningfully contribute to river flow
dynamics. Notably higher Dissolved Oxygen and Salinity are associated
with slower flows, whereas higher pH and Turbidity relate to faster
speeds, highlighting the various complex interactions of physicochemical
properties in shaping river behavior.

Overall, this model offers a robust and interpretable scheme for
understanding river flow, emphasizing that while rivers are inherently
dynamic and influenced by a multifaceted range of external factors,
careful monitoring of these key parameters can provide valuable insights
for environmental management, risk mitigation, and ecological
preservation. 

## Works Cited 

1.  Fain, Jeremy. “Understanding Surface Water Speed in Rivers: Why It
    Matters and How It’s Measured - Blue Water Intelligence (BWI).”
    *Blue Water Intelligence (BWI)*, 8 Sept. 2025,
    <https://bwi.earth/understanding-surface-water-speed-in-rivers-why-it-matters-and-how-its-measured/.>
2.  Fedorov, Daniel. “Water Quality Monitoring Dataset.” *Kaggle.Com*,
    2024,
    <https://www.kaggle.com/datasets/downshift/water-quality-monitoring-dataset.>
3.  Less, Diani F. S., et al. “Seasonal and Daily Variation of
    Hydrodynamic Conditions in the Amazon River Mouth: Influence of
    Discharge and Tide on Flow Velocity.” *Journal of Coastal Research*,
    vol. 37, no. 6, Aug. 2021,
    <https://doi.org/10.2112/jcoastres-d-21-00010.1.>
4.  Trent, Roy, and Scott Brown. *An Overview of Factors Affecting River
    Stability*. 30 Dec. 1984,
    <https://onlinepubs.trb.org/Onlinepubs/trr/1984/950/950v2-017.pdf.>
5.  Queensland Government. “Home." www.data.qld.gov.au, 23 Apr. 2020,
    [www.data.qld.gov.au/.](https://www.data.qld.gov.au/dataset/brisbane-river-colmslie-site-water-quality-monitoring-buoy/resource/0ec4dacc-8e78-4c2a-aa70-d7865ec098e2)
6.  Wickham, Hadley, and Garrett Grolemund. R for Data Science : Import,
    Tidy, Transform, Visualize, and Model Data. O’reilly, 2017,
    <https://r4ds.had.co.nz/index.html.>
